Mass surveillance in cyberspace and the lost art of keeping a secret\thanks{The ideas discussed here first appeared as a DCSS project conference talk and were subsequently fleshed out in paper due to serendipitous encounters fostered by twitter.}

### 2015-07-08 paper.tex
Upload version
w/ Michael's proof reading corrections implemented.

\title{Mass surveillance in cyberspace and the lost art of keeping a secret\thanks{The ideas discussed here first appeared as a DCSS project conference talk and were subsequently fleshed out in paper due to serendipitous encounters fostered by twitter.}}
\subtitle{Policy lessons for nation-states after the Snowden leaks}

@@

Two years after the initial release of Snowden files, surveillance legislation remains highly contested in Canada, the US and the UK. Perhaps most notably is the sunsetting of section 215 of The Patriot Act and subsequent passing of The Freedom Act in the United States in early June~\cite{Patriot}. Days later the Senate of Canada passed controversial anti-terrorism Bill C-51, which received sustained public opposition from big business, journalists, law professors, activists and the privacy commissioner~\cite{C-51}. A week prior to these developments the latest rendition of the so called 'snooper's charter' in the UK was announced in the Queen's speech. Former deputy Prime Minister Nick Clegg publicly opposed the legislation, currently known as The Investigatory Powers Bill, arguing it threatens the privacy rights of citizens. It's details are not fully fledged yet at the time of writing, but it is expected to be a contentious bill and Prime Minister David Cameron has at various points pointed out even towards attempts to curb the use of strong encryption -- albeit there would be fierce opposition to legislation against it.

These measures are indicative state attempts to curb terrorism threats by enabling the development of surveillance capabilities that are of bulk collection nature, rather than targeted to specific individuals. Proponents of these argue that the proliferation of high technology, including anonymity, cryptography and secure communication tools, enables organised crime to communicate safely and go undetected. On the other hand privacy activists advocate the fundamental need for safe spaces to develop one's ideas, the human right to privacy and an individual's need to protect themselves from abusive regimes.

@@

\section{Background}
\label{sec:Background}
The debate on mass surveillance, which is comprised of several threads, has engaged a range of social groups including politicians, law makers, journalists, academics, tech firms, activists, artists and the general public. The term mass surveillance is used to distinguish the bulk collection of data from targeted surveillance, which typically involves a 'person of interest'. Central to this aspect of the debate is the legal warrant, which is traditionally issued upon satisfaction of a certain level of suspicion. In the case of Canada, for example, Bill C-13, which was passed in the fall of 2014, significantly lowered the level of suspicion required to justify the collection of personal data. Bill C-13 also addressed the distinction between data and meta data, which is a hotly debated topic in surveillance legislation. Advocates of expanded surveillance powers for the state have attempted to mollify concerns by arguing that meta data does not threaten the political or civil rights of citizens because it is data about communication and not the content of communication. This argument has been routinely problematized by opponents who point out that metadata can reveal religious beliefs, political leanings and intimate relationships. Moreover, meta data is used by state actors to kill people, as was famously announced by former NSA and CIA director Michael Hayden~\cite{Hayden}.

@@

The Snowden leaks revealed a wide portfolio of projects and initiatives both from the NSA in the US and GCHQ in the UK. These range from specific data collection projects such as Optic Nerve, aimed at Yahoo! webcam traffic (Figure \ref{fig:Yahoo}), to influencing the development of cryptographic standards to contain vulnerabilities, so they can be penetrated easier~\cite{ECDH}. In this varied context the Anderson report~\cite{Anderson} that was released recently as a comprehensive review of the UK's capabilities and practice prior to revamping the existing legislation, emphasised a number of issues, amongst the most important - and contested - of which, was the suggestion for judicial rather than ministerial oversight.

@@

But the 'clinical' perception of the algorithmic component is problematic as well. Just as errors in human judgement may lead to tragic outcomes such as the shooting of Brazilian citizen Charles de Menezes in London by police in the aftermath of the 7/7 attacks, similarly algorithms may equally fail (the headline of Figure \ref{fig:Algo} is indicative of such a failure). In fact the absence of human judgement may make this aspect to be perceived as even more untrustworthy.

@@

Despite the fact that providers of services such as Google and facebook usually operate in multi-jurisdictions, which make difficult a coherent legislative approach and leave a lot of issues with respect to protection of personal data, there is a huge take up of their services. It seems that the personal value realised for each user, combined with the unclear implications/risks to the individual from their use have contributed largely to this. This is despite general concerns of legislators for their operation, as in the case of Belgium that investigated facebook tracking of users, even when not logged in~\cite{facebookBel}. Also despite journalists and researchers have flagged how both in terms of practicalities such as extended data retention periods (e.g.~\cite{Wired}), but also how theoretically can be shown that providers tend to maximise their payoffs when they misuse personal data~\cite{Anastasopoulou}.

We argue that one of the side effects of this is the casualisation of attitudes towards privacy rights. There's a creeping indifference that could develop to passive acceptance through repeated interaction and use of such technologies -- in a way that Giddens describes as routinisation in his theory of structuration~\cite{Giddens}. However, familiarity with people giving up personal data in return for real value needn't necessarily be viewed negative in the context of cyber surveillance. We will argue in the following section that such relationship can be at the centre of the creation of new surveillance systems, built upon consensus where intelligence is necessary and there's clear understanding of its value to all stakeholders.

@@

We referred earlier to CCTV as an example of a surveillance technology that has been relatively successful, from a technology acceptance and use point of view. That is in the Anglo-Saxon world and in particular in the UK, where it is estimated that there were over 4 million cameras in operation, although police investigations suggest that this may be an over-estimation~\cite{HomeOffice}. It is interesting to note that CCTV overall is really a collection of ad hoc installations of security cameras that offer access to footage of varying quality. These include public spaces monitoring as well as private surveillance systems for both corporate and personal property.

In reality it is a system-of-systems that has emerged out of societal consent and adds some value (mostly evidential) to the law enforcement process. It is not standardised in terms of technical configurations or modes of operation, but of course its use is regulated, as personal data protection legislation applies. Thus notices of CCTV enforcement are for example mandatory to be displayed at the physical spaces that are monitored. Operators are also obliged to obfuscate streams that may be intrusive of personal space that could be accidentally included in the footage, e.g. certain frames that capture nearby windows. And in order to build public confidence, several operation centres in cities, such as traffic controllers, would include in their governance structures some involvement of members of the public, or elected local councillors.

Regardless of the heated debate about the effectiveness of CCTV against crime~\cite{Ditton}\cite{Woodhouse}, its proliferation and acceptance in some societies is notable. This is certainly the case in the UK, where CCTV rose to public prominence largely in the 80's. Part of its take up may be due to the campaign against football hooliganism as well as ground safety fears after the tragedy of Hillsborough. Some researchers also suggest that they became instruments of enforcing an image of tough stance against crime by political parties, especially New Labour~\cite{McHill}.

The simplicity of purpose of a video surveillance system and the ability to relate its operation to a societal challenge (in this case football violence and public safety in grounds) established the technology in the collective social mind as something intrusive, yet necessary. This facilitated take up across public spaces of local councils (e.g. car parks), motorways, even on board means of transport, as well as in private business and buildings. Market forces and accessibility to the technology enabled everyone that may had interest in it to install and operate such systems, turning in a way event he general public co-creators of the overall CCTV system-of-systems.

Transferring this into the cyber domain, there is a need for related technology to be viewed as an enabler, as opposed to be demonised and fundamentally mistrusted. Greater transparency of Internet surveillance programmes and sufficient oversight structures would assure the public of the role of technology. Further education and public understanding of surveillance would help. This is not incompatible with the secrecy that security services claim needs to surround their operations; in fact they ought to assume that adversaries are suspicious of their practices. The Panotpticon metaphor needs to be revisited as until now it has blurred the realisation of the need to intervene earlier in the radicalisation lifecycle to debunk their propaganda messages and the appeal of radicalism to young and vulnerable members of society. There is a need for more human-centric intelligence, open source and targeted operations, as opposed to passive monitoring and algorithmic determination implied by the current paradigm. We believe that if the purpose is clear and the community is able to see the value to them, they will even tolerate the trade off of personal data in return for confronting effectively this threat, as suggested even by the commercial experiences discussed earlier.

>>>>

\title{Mass surveillance in cyberspace and the lost art of keeping a secret\thanks{The ideas discussed here first appeared as a DCSS project conference talk and were subsequently fleshed out in the form of a paper due to serendipitous encounters that were lovingly fostered by the social medium of Twitter.}}
\subtitle{Policy lessons for nation-states after the Snowden leaks}

@@

Two years after the initial release of Snowden files, surveillance legislation remains highly contested in Canada, the US and the UK. Perhaps most notably is the sunsetting of section 215 of The Patriot Act and subsequent passing of The Freedom Act in the United States in early June 2015~\cite{Patriot}. Days later the Senate of Canada passed controversial anti-terrorism Bill C-51, which received sustained public opposition from big business, journalists, law professors, activists and the privacy commissioner~\cite{C-51}. A week prior to these developments the latest rendition of the so called 'snooper's charter' in the UK was announced in the Queen's speech. Former deputy Prime Minister Nick Clegg publicly opposed the legislation, currently known as The Investigatory Powers Bill, arguing it threatens the privacy rights of citizens. It's details are not fully fledged yet at the time of writing, but it is expected to be a contentious bill and Prime Minister David Cameron has at various points pointed towards banning the use of strong encryption -- albeit there would be fierce opposition to legislation against it.

These measures are indicative of state attempts to curb terrorism threats by enabling the development of surveillance capabilities that are of bulk collection nature, rather than targeted to specific individuals. Proponents of these argue that the proliferation of high technology, including anonymity, cryptography and secure communication tools, enables organised crime and terrorists, extremists etc. to communicate safely and go undetected. On the other hand privacy activists advocate the fundamental need for safe spaces to develop one's ideas, the human right to privacy and an individual's need to protect themselves from abusive regimes.

@@

\section{Background}
\label{sec:Background}
The debate on mass surveillance, which is comprised of several threads, has engaged a range of social groups including politicians, law makers, journalists, academics, tech firms, activists, artists and the general public. The term mass surveillance is used to distinguish the bulk collection of data from targeted surveillance, which typically involves a 'person of interest'. Central to this aspect of the debate is the legal warrant, which is traditionally issued upon satisfaction of a certain level of suspicion. In the case of Canada, for example, Bill C-13, which was passed in the fall of 2014, significantly lowered the level of suspicion required to justify the collection of personal data. Bill C-13 also addressed the distinction between data and meta data, which is a hotly debated topic in surveillance legislation. In particular, Bill C-13 reduces the limitations on collecting meta data under the guise that it is not intrusive. Advocates of expanded surveillance powers for the state have attempted to mollify concerns by arguing that meta data does not threaten the political or civil rights of citizens because it is data about communication and not the content of communication. This argument has been routinely problematized by opponents who point out that metadata can reveal religious beliefs, political leanings and intimate relationships. Moreover, meta data is used by state actors to kill people, as was famously announced by former NSA and CIA director Michael Hayden~\cite{Hayden}.

@@

The Snowden leaks revealed a wide portfolio of projects and initiatives both from the NSA in the US, CSIS in Canada and GCHQ in the UK. These range from specific data collection projects such as Optic Nerve, aimed at Yahoo! webcam traffic (Figure \ref{fig:Yahoo}), to influencing the development of cryptographic standards to contain vulnerabilities, so they can be penetrated easier~\cite{ECDH}. In this varied context the Anderson report~\cite{Anderson} that was released recently as a comprehensive review of the UK's capabilities and practice prior to revamping the existing legislation, emphasised a number of issues, amongst the most important - and contested - of which, was the suggestion for judicial rather than ministerial oversight.

@@

But the 'clinical' perception of the algorithmic component is problematic as well. Just as errors in human judgement may lead to tragic outcomes such as the shooting of Brazilian citizen Charles de Menezes in London by police in the aftermath of the 7/7 attacks, similarly algorithms may equally fail (the headline of Figure \ref{fig:Algo} is indicative of such a failure). In fact the absence of human judgement may cause this aspect to be perceived as even more untrustworthy.

@@

Despite the fact that providers of services such as Google and Facebook usually operate in multi-jurisdictions, which make difficult a coherent legislative approach and leave a lot of issues with respect to protection of personal data, there is a huge take up of their services. It seems that the personal value realised for each user, combined with the unclear implications/risks to the individual from their use have contributed largely to this. This is despite general concerns of legislators for their operation, as in the case of Belgium that investigated Facebook tracking of users, even when not logged in~\cite{facebookBel}. Also despite journalists and researchers have flagged how both in terms of practicalities such as extended data retention periods (e.g.~\cite{Wired}), but also how theoretically can be shown that providers tend to maximise their payoffs when they misuse personal data~\cite{Anastasopoulou}.

We argue that one of the side effects of this is the casualisation of attitudes towards privacy rights. There's a creeping indifference that could develop to passive acceptance through repeated interaction and use of such technologies -- in a way that Giddens describes as routinisation in his theory of structuration~\cite{Giddens}. However, familiarity with people giving up personal data in return for real value needn't necessarily be viewed negatively in the context of cyber surveillance. We will argue in the following section that such relationship can be at the centre of the creation of new surveillance systems, built upon consensus where intelligence is necessary and there is clear understanding of its value to all stakeholders.

@@

We referred earlier to CCTV as an example of a surveillance technology that has been relatively successful, from a technology acceptance and use point of view. That is particularly apparent in the UK, where it is estimated that there were over 4 million cameras in operation, although police investigations suggest that this may be an over-estimation~\cite{HomeOffice}. It is interesting to note that CCTV overall is really a collection of ad hoc installations of security cameras that offer access to footage of varying quality. These include public spaces monitoring as well as private surveillance systems for both corporate and personal property.

In reality CCTV is a system-of-systems that has emerged out of societal consent and adds some value (mostly evidential) to the law enforcement process. It is not standardised in terms of technical configurations or modes of operation, but of course its use is regulated, as personal data protection legislation applies. Notices of CCTV enforcement are for example mandatory to be displayed at the physical spaces that are monitored. Operators are also obliged to obfuscate streams that may be intrusive of personal space that could be accidentally included in the footage, e.g. certain frames that capture nearby windows. In Canada individuals may also be blocked out if they are in the frame and not directly relevant to an investigation. And in order to build public confidence, several operation centres in cities, such as traffic controllers, would include in their governance structures some involvement of members of the public, or elected local councillors.

Regardless of the heated debate about the effectiveness of CCTV against crime~\cite{Ditton}\cite{Woodhouse}, its proliferation and acceptance in some societies is notable. This is certainly the case in the UK, where CCTV rose to public prominence largely in the 80's. Part of its take up may be due to the campaign against football hooliganism as well as ground safety fears after the tragedy of Hillsborough. Some researchers also suggest that they became instruments of enforcing an image of a tough stance against crime by political parties, especially New Labour~\cite{McHill}.

The simplicity of purpose of a video surveillance system and the ability to relate its operation to a societal challenge (in this case football violence and public safety in grounds) established the technology in the collective social mind as something intrusive, yet necessary. This facilitated take up across public spaces of local councils (e.g. car parks), motorways, even on board means of transport, as well as in private business and buildings. Market forces and accessibility to the technology enabled everyone that may have had interest in it to install and operate such systems, creating space for the general public to serve as co-creators of the overall CCTV system-of-systems.

Translating this into the cyber domain, there is a need for related technology to be viewed as an enabler, as opposed to being demonised and fundamentally mistrusted. Greater transparency of Internet surveillance programmes and sufficient oversight structures would assure the public of the role of technology. Further education and public understanding of surveillance would help. This is not incompatible with the secrecy that security services claim must surround their operations; in fact they ought to assume that adversaries are suspicious of their practices. The Panotpticon metaphor needs to be revisited as until now it has blurred the realisation of the need to intervene earlier in the radicalisation lifecycle to debunk their propaganda messages and the appeal of radicalism to young and vulnerable members of society. There is a need for more human-centric intelligence, open source and targeted operations, as opposed to passive monitoring and algorithmic determination implied by the current paradigm. We believe that if the purpose is clear and the community is able to see the value to them, they will even tolerate the trade off of personal data in return for confronting effectively this threat, as suggested even by the commercial experiences discussed earlier.
